---
title: "Car Thefts Significance"
author: "Derek Willis/Rob Wells"
date: 10-27-2025
---

name: Riley Bomgardner


Your editor wants to know if there have been recent changes in the number of car thefts in Prince George's County. Using monthly figures and significance testing, you'll come up with an answer and decide whether there's a story here. As with the previous homework, provide answers in this notebook when prompted and you will have to fill in values where you see REPLACE_ME. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# LOAD THE TIDYVERSE
library(tidyverse)
```


### Task 2: Load the Prince George's county crime data and generate car thefts totals by month

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv")  |> 
  arrange(month)  # Sort by date

# Basic inspection of the data
glimpse(car_thefts)

```

### Task 3: Let's make a chart and describe the results.

We'll make a line chart using geom_line() to try and illustrate any trend over time. 

```{r}

car_thefts |> 
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Car Thefts (July 2023 - February 2025)",
    subtitle = "Is there a trend over time?",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

```


**Q3:** What's your initial reaction to the change in car thefts? Is there a trend?

**WRITE ANSWERS HERE:** I would say no. Over time, if you look at the value for car thefts in July of 2023, it's a much higher number than February of 2025, so there is a change. But there are a lot of fluctuations throughout the graph that make it inconsistent and harder to define as a trend. All the values after July of 2023 are less than the 673 thefts in that month, so there is a obvious drop.


### Task 4: Add the mean, standard error and calculate a margin of error based on a 95% confidence interval

Let's add some stats to that process - mean and standard deviation - and calculate the standard error and 95% confidence intervals. We only have a small number of months, so we'll calculate the margin of error to account for that. 

**INSERT CORRECT VARIABLE AT "REPLACE ME"**

```{r}

theft_summary <- car_thefts |> 
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts,  # Using t-distribution for small sample
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

# Display the confidence interval
theft_summary
```




### Task 5: Visualize the data

Let's make a better chart showing the changes and the statistics we produced in Task 4. Does this change anything about your answer from Task 3?



```{r}
# Make a better chart
car_thefts |> 
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 2) +
  geom_hline(yintercept = theft_summary$mean_thefts, linetype = "dashed", color = "red", size = 1) +
  geom_ribbon(aes(ymin = theft_summary$ci_lower, ymax = theft_summary$ci_upper), 
              fill = "red", alpha = 0.2) +
  annotate("text", x = min(car_thefts$month), 
           y = theft_summary$mean_thefts + 10, 
           label = paste0("Mean: ", round(theft_summary$mean_thefts, 1)), 
           hjust = 0, color = "red") +
  labs(
    title = "Monthly Car Thefts with 95% Confidence Interval for the Mean",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal()

```

### Task 6: Describe the confidence interval.

- What does the 95% confidence interval show here? **WRITE ANSWER HERE: The 95% confidence level tells us what range we can be 95% sure the true mean for Prince George's County monthly car thefts can fall under. This range looks to be from about 420 to 510 car thefts per month, meaning more than half of the monthly figures should fall within this range, as seen with only a few outliers above and below the red ribbon on the graph.**

- What figure(s) accurately reflects the number of car thefts in a typical month? **WRITE ANSWER HERE: The mean thefts per month, along with the margin of error. The values from the range of the confidence interval are about the +/- 45.26 margin of error.**

- Propose a headline about the trend in car thefts that is focused on the present and future, but informed by the data you see above. **WRITE ANSWER HERE: Car thefts in Prince George's County drop from July 2023 high, but still fluctuate on a month-to-month basis.**



### Task 7: Is the recent trend significant?

It _looks_ like car thefts are decreasing lately. But is that change real? We'll split the monthly data into two halves, and then ask if whether there's a statistically significant difference between these periods. To do that, we'll use a t-test with our alpha at 0.05 and develop a null hypothesis and an alternative hypothesis for this scenario.

Null Hypothesis: there is no real difference between the two periods
Alternative Hypothesis: there is a real difference between the two periods

**INSERT CORRECT VARIABLES AT "REPLACE ME"**

```{r}

# Find the halfway point and add a column placing each month in the first or second halves

midpoint <- ceiling(nrow(car_thefts) / 2)
car_thefts <- car_thefts |> 
  mutate(period = if_else(row_number() <= midpoint, "First Half", "Second Half"))

# Perform t-test comparing the two time periods - this is a two-sided test because we have two periods
t_test_periods <- t.test(total ~ period, data = car_thefts)

# Display results
t_test_periods

# Visualize the comparison
period_summary <- car_thefts |> 
  group_by(period) |> 
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts, # look at how we calculated this above
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

period_summary |> 
ggplot(aes(x = period, y = mean_thefts, fill = period)) +
  geom_col() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_text(aes(label = round(mean_thefts, 1)), vjust = -0.5, size = 4) +
  labs(
    title = "Comparing Car Thefts: First Half vs. Second Half of Data Period",
    subtitle = paste0("p-value = ", round(t_test_periods$p.value, 3)),
    x = "Time Period",
    y = "Average Number of Car Thefts"
  ) +
  scale_fill_manual(values = c("First Half" = "#66c2a5", "Second Half" = "#fc8d62")) +
  theme_minimal() +
  theme(legend.position = "none")

```

### Task 8: Reflection

Given the above chart, is the difference between the two periods statistically significant? Can we reject the null hypothesis? **WRITE ANSWER HERE: We cannot reject the null hypothesis because the p-value (0.179) is greater than the 0.05 cutoff, meaning the difference between the two periods is not statistically significant. Even though the second half of the data is lower in average (430.8) than the first half (490), this could just be due to variation and not a downwards trend.**

How would you revise your headline from Task 6, if at all? What would you tell your editor, and what would you expect her to ask you? **WRITE ANSWER HERE: Car thefts show a slight decline in Prince George's County, but there's no statistical significance.   I would tell my editor about the t-test that was run and explain what the higher p-value means in relation to the cutoff and why the data isn't statistically significant. I would also bring up some examples as to why this certain section of crime has seen a drop, like bringing up other forms of transportation that may have become more accessible in the 20 months this data covers (buses, metro, e-scooters, etc.) I would also see if there can be any comparison among other counties in Maryland to see if crime is decreasing across the state or just in PG.**





When you are finished, add, commit and push your notebook and submit the URL in ELMS.
