---
title: "HW12 car theft anova"
author: "Your Name"
---

**Riley Bomgardner**


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Introduction

This analysis examines the impact of a significant spike in car theft rates in 2023 in Prince George's County, and whether that spike has led to a sustained higher rate of car theft in the county. We'll use Analysis of Variance (ANOVA) to determine if theft rates differ significantly between pre-spike (before 2023), spike (2023), and post-spike (2024-present) periods.

## Data Preparation and Exploration

### Loading Libraries and Data

```{r load-libraries}
# Load required libraries
library(tidyverse)
library(lubridate)
```

```{r load-data}
# Import data
theft_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month17_25.csv")

# Look at the first few rows
head(theft_data)

# Get a summary of the data
summary(theft_data)
```


### Get the data how we want it and calculate the rates

**Task 1** Replace REPLACE_ME with the variables you need to calculate the theft_rate (2 points)

```{r process-data}
# Calculate theft rate and create time period variable
theft_data <- theft_data |>
  mutate(
    date = as.Date(month),
    year = year(date),
    month_number = month(date),
    theft_rate = (total / population) * 100000,
    time_period = case_when(
      year < 2023 ~ "Pre-spike",
      year == 2023 ~ "Spike year",
      year > 2023 ~ "Post-spike"
    )
  )

# Create summary by time period
period_summary <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_thefts = mean(total),
    mean_rate = mean(theft_rate),
    median_rate = median(theft_rate),
    sd_rate = sd(theft_rate),
    n = n(),
    .groups = "drop"
  )

# Display the summary
period_summary
```

**Task 2** Describe the results of calculating the rates for each period. What do they suggest? (3 points)

**WRITE YOUR ANSWER HERE**: The mean and median rates almost triple from pre-spike (23.19 mean and 22.17 median) to spike year (59.95 mean and 58.13 median). The data also shows that mean thefts still have a higher value post-spike (418.79) than pre-spike (216.42), meaning that it's taking time to get car thefts consistent again.

## Visualizing the Spike Pattern

### Time Series Plot

```{r time-series}
# Time series plot showing the spike pattern
ggplot(theft_data, aes(x = date, y = theft_rate)) +
  geom_line() +
  geom_smooth(method = "loess", span = 0.2) +
  geom_vline(xintercept = as.Date("2023-01-01"), linetype = "dashed", color = "red") +
  geom_vline(xintercept = as.Date("2024-01-01"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("2023-07-01"), y = max(theft_data$theft_rate) * 0.9, 
           label = "Spike Year", color = "red") +
  labs(title = "Monthly Car Theft Rates (2017-2025) with Spike Highlighted",
       x = "Date", y = "Thefts per 100,000 Population") +
  theme_minimal()
```

**Task 3** Describe the output of this chart and then give it a better, more active title that provides the lede (3 points)

**WRITE YOUR ANSWER HERE**: This chart shows all car thefts in Prince George's County from 2017 through 2024. The spike year is highlighted by red dashed lines, showing enormous increase in thefts at the end of 2022 through the end of 2023. You can also see where the decline after the spike begins, but it's no where near the values of the pre-spike data.

Car thefts soared in 2023 in Prince George's County, but continue to stay elevated post-spike.


Now let's take a look at average theft rates.

### Yearly Average Theft Rates

```{r yearly-means}
# Calculate yearly means
yearly_means <- theft_data |>
  group_by(year) |>
  summarize(
    mean_rate = mean(theft_rate),
    .groups = "drop"
  )

# Plot yearly means
ggplot(yearly_means, aes(x = year, y = mean_rate)) +
  geom_line(size = 1) +
  geom_point(size = 3, color = "blue") +
  geom_point(data = yearly_means |> filter(year == 2023), 
             aes(x = year, y = mean_rate), size = 4, color = "red") +
  labs(title = "Average Annual Car Theft Rates",
       x = "Year", y = "Average Thefts per 100,000 Population") +
  theme_minimal() +
  scale_x_continuous(breaks = 2017:2025)
```

## One-way ANOVA Analysis

### Hypothesis Formulation

For our ANOVA test, we'll compare the mean theft rates between the three time periods. First we need to create our hypotheses.

**Task 4** Fill in the hypotheses (2 points)

- **Null Hypothesis (H0)**:   There is no significant difference between mean theft rates of pre-spike, spike, and post-spike.

- **Alternative Hypothesis (H1)**: There is a significant difference between mean theft rates of pre-spike, spike, post-spike.

### One-way ANOVA Test

```{r anova-test}
# Perform one-way ANOVA on theft rates by time period
period_anova <- aov(theft_rate ~ time_period, data = theft_data)
summary(period_anova)
```

**Task 5** Describe the result: are there real differences in the average rates between the time periods? Is this result significant? (5 points)

**WRITE YOUR ANSWER HERE** Yes, there are real differences. This test confirms that there is a statistical significance because the p-values is less than 0.001. If it were higher, we would have to accept the null hypothesis and say there is not a significant difference in time periods.

## Compare the differences between specific periods

Run our post-anova Tukey test and examine the results. 

```{r posthoc}
# Tukey's HSD test
tukey_results <- TukeyHSD(period_anova)
print(tukey_results)
```


```{r period-means}
# Visualize time period means with confidence intervals
period_rates <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_rate = mean(theft_rate),
    se = sd(theft_rate) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(
    lower_ci = mean_rate - qt(0.975, n() - 1) * se,
    upper_ci = mean_rate + qt(0.975, n() - 1) * se
  )

# Ensure time period is ordered correctly
period_rates$time_period <- factor(period_rates$time_period, 
                                   levels = c("Pre-spike", "Spike year", "Post-spike"))

ggplot(period_rates, aes(x = time_period, y = mean_rate, fill = time_period)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  labs(title = "Car Theft Rates by Time Period with 95% Confidence Intervals",
       x = "", y = "Mean Theft Rate per 100,000") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Conclusion

This analysis has examined whether there are significant differences in car theft rates between the pre-spike period (before 2023), the spike year (2023), and the post-spike period (after 2023).

**Task 6** Describe the results of the Tukey's test and chart showing confidence intervals. Is there a significant difference between the three periods? In particular, is the difference between the pre-spike and post-spike periods real and significant? Write a headline and lede reporting these findings, focusing on the pre- and post-spike periods. (5 points)

**WRITE YOUR ANSWER HERE**: The Tukey shows the significant differences in each mean car theft rate for pre-spike, spike, and post-spike. There is a significant different between pre and post-spike, but it's smaller than if compared to the spike year. Theft rates didn't return to pre-spike levels, which is what the chart shows us. You can see how significant the difference is just by looking at the gaps on the chart. The confidence levels for pre-spike and post also do not overlap, so there is significance to this data.

Two years after a 2023 spike, car thefts remain high in Prince George's County 

Although car thefts rates in Prince George's County have declined from a 2023 spike, they remain well above pre-spike levels, new data shows. Statistical tests run by journalism students at the University of Maryland confirm that car thefts haven't decreased all the way back to pre-spike numbers, highlighting that the spike from 2023 continues to have lasting effects on car-related crime in the county.