---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse, janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities %>%
  mutate(
    violent_rate_2019 = (violent2019 / pop2019) * 100000,
    property_rate_2019 = (property2019 / pop2019) * 100000,
    violent_rate_2020 = (violent2020 / pop2020) * 100000,
    property_rate_2020 = (property2020 / pop2020) * 100000
  )
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

ANSWER HERE: Ocean City in Worcester has the highest property crime for both 2019 and 2020, even though its population is not very large. Their violent rate also increases from about 1310 per 100000 in 2019 to almost 2400 per 100000 in 2020, which is the highest violent rate for the cities in Maryland in 2020. Many of the safest cities are located in Allegany County (Lonaconing, Midland, and Westernport), recording a 0 in all categories.


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates %>%
  mutate(
    violent_pct_change = ((violent_rate_2020 - violent_rate_2019) / violent_rate_2019) * 100,
    property_pct_change = ((property_rate_2020 - property_rate_2019) / property_rate_2019) * 100
  )
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

ANSWER HERE: For the most part, the property rates percentage decreased, with the exception of some cities who had no crime over the two years or those with a dramatic increase in property crimes from 2019 to 2020. The highest percentage change in violent crimes was 895 in Edmonston in Prince George's County. This is significantly higher than other percentage changes, except for Poolesville in Montgomery County, which recorded a 607 percent change. Something that stood out to me was that Baltimore's violent crime percentage change decreased by around 12%. This is most likely due to Covid happening, but I wonder if there were other factors.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE: Most of Maryland's safest cities in 2019 and 2020 are in western counties and the Eastern Shore, according to crime and property rates given for cities in each county. 

These "safe cities" also share something else in common: Their populations are all under 2000 people for both years. Factor in that the vast majority of these cities are surrounded by mountains or farmland. More distance means less crime, and can also mean less is reported if the distance to police stations is large and time-consuming. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.


HW5: PERCENTILES


### Task 10. Your task will be to take the violent crime rates for 2020 and do three things:

Using the 2020 violent crime rate, calculate the percentiles using `ntile()`. Save the results to a new dataframe called `md_cities_percentiles`. How many cities are above the 90th percentile?
```{r}
md_cities_percentiles <- md_cities_change %>%
  mutate(violent_percentile_2020 = ntile(violent_rate_2020, 100)
  )
```

How many cities are above the 90th percentile?

ANSWER HERE: There are 10 cities above the 90th percentile: Clear Spring (Washington), Salisbury (Wicomico), Seat Pleasant (PG), Elkton (Cecil), Williamsport (Washington), Kitzmiller (Garrett), Princess Anne (Somerset), Cambridge (Dorchester), Baltimore City and Ocean City (Worcester).


Calculate the Interquartile Range of the 2020 violent crime rate using `summarize()` and R's `IQR()` function. The syntax is similar to how `mean()` works.
```{r}
md_cities_change %>%
  summarize(
    violent_IQR_2020 = IQR(violent_rate_2020, na.rm = TRUE)
  )
```
Do the results change your best story idea from the previous assignment? Why or why not?

ANSWER HERE: Yes, because many of the cities above the 90th percentile in violent crimes are located on the Eastern Shore and the western counties of Maryland, which is the opposite of what my story was about. Ocean City has the highest number of violent crimes percentile in the state (100), beating Baltimore! My story would have to be changed, and it would most likely be asking about why the violent percentile especially is so high on the east and west sides of the state, but not very much in the middle.

Rename the notebook as HW5_Percentiles, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.