---
title: "final_project"
output: html_document
date: "2025-12-15"
Riley Bomgardner
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse, janitor)
```

**Final Project: Billboard Hot 100**
What makes a hit song? Is it rhythm, lyrics, popularity of the artist, beats per minute? Does the day a song is released on have any impact on if it's a hit? How does BPM influence danceability and therefore the ability of a song to become a hit? That's what will be tested in this final project.

```{r load data}
billboard_100 <- read_csv("https://raw.githubusercontent.com/rbomgard-creator/jour405/refs/heads/main/project/BillboardHot100NumberOnesDatabaseData.csv")

# preview structure
view(billboard_100)
```



**Notes**

1. link to data source and details on the data: https://docs.google.com/spreadsheets/d/1MSp-Feo0X54bthtB5KT5ucSzUEwS89DHqVgKqDVOC7g/edit?gid=1974823090#gid=1974823090

2. More recent songs (Old Town Road (2019) and A Bar Song (Tipsy) (2024)) have the highest number of weeks at number 1 (19 weeks). Is a song less of a hit if it only stays on the hot 100 for a certain amount of weeks? Is a song not a hit if it's only on the charts for 1 week?

3. How is billboard's "rating" measured? How important is this number to establishing if one song is more of a hit than another?

4. Covers the weekly Billboard Hot 100 from August 4, 1958 to January 11, 2025.

5. Important columns: Weeks at Number 1, danceability, energy, BPM, time signature, CDR Genre, Discogs Genre, Overall Rating, Date, Song, Artist

6. Billboard Hot 100s are released on Mondays


**Statistics on time signature**
```{r}
billboard_100 |>
  summarize(
    mean_time_sig = mean(`Time Signature`, na.rm = TRUE),
    median_time_sig = median(`Time Signature`, na.rm = TRUE),
    min_time_sig = min(`Time Signature`, na.rm = TRUE),
    max_sig = max(`Time Signature`, na.rm = TRUE),
    sd_sig = sd(`Time Signature`, na.rm = TRUE)
  )
```
Probably not the best tests to run on this type of measurement, but it seems that the 4/4 time signature is used quite often among these billboard hits. None of these time signatures are anything that would be rare, as they are used in any type of music you can think of. It may be more of a BPM measure that truly determines if a song is a hit, but it needs a time signature to go along with it.



**Stats on BPM**
```{r}
billboard_100 |>
  summarize(
    mean_bpm = mean(BPM, na.rm = TRUE),
    median_bpm = median(BPM, na.rm = TRUE),
    min_bpm = min(BPM, na.rm = TRUE),
    max_bpm = max(BPM, na.rm = TRUE),
    sd_bpm = sd(BPM, na.rm = TRUE)
  )
```
115.85 BPM is the mean value for this test. This is almost two beats per second. This means that most hits share similar beats per minute, which could be one of the main reasons these songs reach such charts. This should be compared to the overall rate as well, but the date for these songs is important because it could determine what people determined as a hit in other decades. For example, "Every Breath You Take," by The Police is listed at 117 beats per minute, but gives the impression of a slower song because of its mellow tone. Does tone determine a hit? What about the song's message? All things to keep stats on.

It may be a good idea to divide the music by decade at a later time just to see if there's any difference in bpm or overall rate, or how long a song has stayed at number one on the charts. Or if different genres are more of a hit over one decade or another (hip-hop in the 90s may produce more hits than hip-hop in the 70s).


**Hypothesis**

Null hypothesis: Any song can be a hit and beats per minute doesn't factor into that. (no significance)
Alternative Hypothesis: If a song has a BPM more than or around 115, it's likely to become a hit and be part of the Billboard Hot 100. (significance)

How much does BPM affect the charts? I think something else that might be nice to include in the data would be the shift from physical music (cassettes, vinyls, radios/cds) to more streaming sites (Spotify, Apple Music, Amazon Music, etc) because I feel like this could affect popularity of a song. Like is the Billboard Hot 100 based on streaming now, or still on what's played on the radio? Is it a mix of both?
How do sales come into play? Are these songs hits because of a high revenue in sales? Are there songs on this list that may have lower revenues but are still considered a hit?

I think the shift in how music is listened to should be noteworthy because it may have been weighed differently before streaming came into the picture.




Saved to GitHub as of 11:32 p.m. on November 17th, 2025.




**Part 2: Running Tests**
First, fix the date format.
```{r}
billboard <- read_csv("https://raw.githubusercontent.com/rbomgard-creator/jour405/refs/heads/main/project/BillboardHot100NumberOnesDatabaseData.csv") |> 
clean_names() |> 
mutate(date_new = dmy(date)) |> 
mutate(date_new = if_else(year(date_new) > 2025,  date_new - years(100), date_new)) |> 
select(date_new, everything())
```
Now, I have a new table with a date format this is easier to use in analysis and further testing. 

As mentioned in the first part of the final project, I'm testing a null hypothesis that states the BPM has nothing to do with a song becoming a hit, while an alternative hypothesis states otherwise. 

This needs to change a bit since I have nothing to compare it too. Instead, does BPM have an impact on how long a hit stays at #1? That's what were testing.

Null: BPM does not affect how long a hit stays on the charts
Hypothesis: Songs with a higher BPM (greater or equal to 115) stay at #1 for a different amount of time than those with lower BPMs. 

To test the significance of an alternative hypothesis, I'm going to run an t-test test, since I'm only using BPM.
```{r}
# T-Test: Do songs with BPM >= 115 stay at #1 longer?

# Create binary variable for high BPM (>= 115)
billboard_high_bpm <- billboard |>
  filter(!is.na(bpm) & !is.na(weeks_at_number_one)) |>
  mutate(high_bpm = bpm >= 115)

# Run the t-test
bpm_ttest <- t.test(weeks_at_number_one ~ high_bpm, data = billboard_high_bpm)

# Display results
print(bpm_ttest)
```
The null hypothesis can be excepted in this instance. The p-value came out to be 0.80, which is way over the significance value of 0.05. There is no statistically difference in the number of weeks at #1 between songs with a BPM greater than or equal to 115 and those with a BPM less than 115. This test also shows that songs less than the 115 average tend to stay at #1 for an average of 2.92 weeks, while those equal or greater than a BPM of 115 tend to stay at number one for around 2.96 weeks. This difference is basically nothing, which makes it statistically insignificant.

Next, I think it would be wise to see if decade affects BPM. Is there a drastic change among decades or does it mainly stay the same? 
Null: There is no significant difference in mean BPM across decades.
Alternative: At least one decade has a significantly different mean BPM compared to other decades. 

Let's run an ANOVA test.
```{r}
# Create decade variable from date
billboard_decades <- billboard |>
  filter(!is.na(bpm) & !is.na(date_new)) |>
  mutate(
    year = year(date_new),
    decade = floor(year / 10) * 10,
    decade_label = paste0(decade, "s")
  )

# Check decade distribution
decade_counts <- billboard_decades |>
  count(decade_label, sort = TRUE)

print("Songs per decade:")
print(decade_counts)

#ANOVA test
anova_result <- aov(bpm ~ decade_label, data = billboard_decades)
anova_summary <- summary(anova_result)

print("ANOVA Results:")
print(anova_summary)
```
The null hypothesis can also be accepted here. The p-value is 0.067, which is very close the significance level, but not less than. So there is not a significant difference in BPM for each decade. However, because this is so close to the cut-off, you may wonder if some decades might have a trend in BPM, but the evidence and ANOVA test isn't strong enough to reach an absolute conclusion. 

Now, a data visualization. I've decided to keep the decade and BPM factors for this graph. 
```{r}
billboard_decades |> 
  group_by(decade_label) |>
  summarize(mean_bpm = mean(bpm, na.rm = TRUE)) |>
  ggplot(aes(x = decade_label, y = mean_bpm)) +
  geom_line(aes(group = 1), color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Mean BPM by Decade (1950s - 2020s)",
    subtitle = "Is there a trend over time?",
    x = "Decade",
    y = "Average Beats Per Minute (BPM)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
This chart tells me that there is no real trend over time, even though it seems like the average BPM has increased close to linearly since the 90s. However, there are fluctuations, like from the 80s to the 80s that create inconsistencies, so this graph tells me further why I had to accept the null hypothesis in the ANOVA test above. If I was to only test from the 90s to now, the statistics may actually be significant, but I needed to measure the full list of dates given to me. 
The 50s and 80s have similar BPMs. One similar BPM is 113, which showcases "Smoke Gets in Your Eyes," from the Platters (1959), and "West End Girls," from Pet Shop Boys (1986). Although the 80s introduce "new wave" as a newer genre, these songs share a BPM and are both catchy in their own ways. 
You can also take into account a cultural context of these songs. During the 90s, grunge, alternative, and slower R&B ruled the Billboard Hits lists, while the 2000s-2020s accelerated with EDM influences, TikTok viral songs, etc. Modern songs are faster, but due to analyzing the entire chart, this becomes less significant. 


Saved to GitHub as of 11:49 pm, November 25, 2025.



***Part 3: Story Pitch***

Working Headline:

Does Tempo Really Make a Hit? What Billboard’s No. 1 Songs Reveal About BPM and Popularity


Story Pitch:

The idea of a magic formula for a hit song has long been the subject of speculation among music industry insiders and listeners. One common belief is that more energetic songs - measured by beats per minute - are both more likely to reach No. 1 on Billboard Hot 100 and then stay on the charts longer. Using Billboard Hot 100 No. 1 data from 1958 through early 2025, this analysis tested if BPM does indeed play a significant role in determining how long a song stays at No. 1.
It would appear that tempo is not, on its own, the chief determining factor for chart longevity. Songs with higher BPMs - most commonly defined here as being 115 BPM or above - did not stay at No. 1 significantly longer than slower songs. A t-test showed nearly identical averages - about three weeks at No. 1 for both groups - with no statistical significance. Furthermore, an additional ANOVA test found no significant variation in average BPM across decades - perhaps surprising given that many people feel that modern hits are faster than older ones.
This conclusion challenges the idea that faster music is inherently more successful on the charts and suggests that other factors such as genre trends, cultural context, marketing, and changes in how music is consumed play a larger role than tempo.


Key Highlights from Findings:

The average BPM of Billboard No. 1 hits is around 116 BPM, a tempo that has remained pretty consistent through time.
Songs with BPMs above this average do not stay at No. 1 longer than slower songs.
While BPM does seem to have a slight uptick in the last few decades of the chart, the difference is not quite statistically significant over the full Billboard chart history.
Even with changing listening habits and platforms, tempo alone cannot explain success with regards to musical success.


Limitations and Potential Problems:

The present analysis only includes songs that reached No. 1, meaning it cannot compare hit songs to non-hits. As a result of that, BPM's role in becoming a hit cannot be fully tested - only its relationship to chart longevity. Moreover, Billboard's ranking methodology changed multiple times over the 1958-2025 timeline, especially with the introduction of streaming and digital sales. It also changed how success is measured across eras. Other influential variables include marketing budgets, radio airplay, genre popularity, and artist recognition, which are not directly measured in this dataset.


Reporting Plan:

To strengthen the story, reporting would involve:
--Interviews with experts in the music industry, producers, or musicologists on whether BPM is consciously considered when composing songs.
--Additional context from Billboard editors or other music analysts about how the calculations for the charts have changed with streaming and social media.
--Examples of iconic hits that had slower tempos and thus run opposite to what common assumptions about energy and popularity.
--A comparison between the BPM trends and genre shifts over time, in order to better explain cultural influences on chart success.


All in all, the data and reporting paint a nuanced picture: Whereas tempo does matter to how music feels, it's not a great predictor of chart dominance. The factors that have decided success on the Billboard Hot 100 seem to be more influenced by cultural moments and industry forces rather than just beats per minute.


Saved to Github as of 11:34pm, Dec. 12, 2025.
